---
title: "Getting started with killerPals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with killerPals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  dpi = 120
)
```

```{r libs}
library(killerPals)
library(ggplot2)
```

killerPals ships 48 palettes: a qualitative, a sequential and a diverging
palette for each of Queen's fifteen studio albums, plus a `greatest_hits` family
drawn from the whole catalogue.

```{r names}
killer_names()
```

## Picking a palette type

The single most common mistake in colour choice is using the wrong *type* of
palette for the data. killerPals makes the three types explicit:

| Your variable | Palette type | Scale |
|---|---|---|
| Unordered categories (species, country, treatment) | `qualitative` | `scale_*_killer_d()` |
| Ordered, low to high (count, density, temperature) | `sequential` | `scale_*_killer_c()` |
| Spread either side of a meaningful zero (change, residual, anomaly) | `diverging` | `scale_*_killer_c(type = "diverging")` |

### Qualitative

```{r qual}
ggplot(mpg, aes(class, fill = class)) +
  geom_bar() +
  scale_fill_killer_d("all_that_jazz") +
  theme_minimal() +
  theme(legend.position = "none")
```

Qualitative palettes deliberately refuse to invent extra colours. Asking for
more than the palette defines is an error, not a silent interpolation, because
interpolating between colourblind-safe colours produces colours that are not:

```{r qual-limit, error = TRUE}
killer_pal("flash", n = 12)
```

If you genuinely have a dozen categories, use `greatest_hits` — the largest
qualitative palette — or reconsider whether twelve colours will communicate
anything at all.

```{r qual-n}
length(killer_pal("greatest_hits"))
```

### Sequential

```{r seq}
ggplot(faithfuld, aes(waiting, eruptions, fill = density)) +
  geom_raster() +
  scale_fill_killer_c("heavenly") +
  theme_minimal()
```

Every sequential ramp is strictly monotone in lightness, which is what makes it
readable in greyscale and for readers with any kind of colour vision:

```{r seq-mono}
killer_check("heavenly", n = 9, type = "sequential")$monotone_lightness
```

### Diverging

A diverging palette says "this value is above or below a reference point". That
only works if the palette's light midpoint actually sits at that reference
point, so **set symmetric `limits`**:

```{r div}
set.seed(1975)
d <- expand.grid(x = 1:24, y = 1:24)
d$z <- as.vector(scale(outer(1:24, 1:24, function(a, b) sin(a / 4) * cos(b / 5))))

ggplot(d, aes(x, y, fill = z)) +
  geom_raster() +
  scale_fill_killer_c("kind_of_magic", type = "diverging",
                      limits = c(-2.5, 2.5)) +
  theme_minimal()
```

Leave the limits off and killerPals will remind you, because the default range
is whatever your data happens to span — which will silently put the midpoint
somewhere meaningless:

```{r div-warn}
p <- ggplot(d, aes(x, y, fill = z)) +
  geom_raster() +
  scale_fill_killer_c("kind_of_magic", type = "diverging")
```

### Binned

Binned scales cut a continuous variable into discrete steps, which often reads
more clearly than a smooth gradient:

```{r binned}
ggplot(faithfuld, aes(waiting, eruptions, fill = density)) +
  geom_raster() +
  scale_fill_killer_b("robot_news", n.breaks = 6) +
  theme_minimal()
```

## Using colours outside ggplot2

`killer_pal()` returns plain hex strings, so the palettes work anywhere.

```{r base}
pal <- killer_pal("hot_space", n = 4)
pal

barplot(1:4, col = pal, border = NA, names.arg = LETTERS[1:4])
```

Sequential and diverging palettes interpolate through CIELAB to any length you
ask for:

```{r interp}
killer_pal("miracle", n = 3, type = "sequential")
killer_pal("miracle", n = 11, type = "sequential")
```

For code that needs a palette *function* — the shape ggplot2 and scales expect —
use `killer_pal_fun()`:

```{r palfun}
f <- killer_pal_fun("innuendo", type = "diverging")
f(5)
```

## Auditing accessibility

The package's central claim is that these palettes stay legible for colourblind
readers. `killer_check()` lets you verify that rather than trust it.

```{r check}
killer_check("greatest_hits")
```

`min_distance` is the smallest [CIEDE2000](https://doi.org/10.1002/col.20070)
difference between *any two* colours in the palette, recomputed after simulating
each kind of colour vision deficiency. As a rule of thumb, a difference of 1 is
the smallest a person can detect at all, and 10 is comfortably distinguishable
at a glance.

To see it rather than read it:

```{r cvd-grid, fig.height = 3}
killer_cvd_grid("greatest_hits")
```

Compare that with a palette built without any of this in mind:

```{r rainbow, fig.height = 2.4}
rb <- grDevices::rainbow(8)
op <- par(mar = c(0.5, 5.5, 1.5, 0.5))
plot(c(0, 8), c(0, 2), type = "n", axes = FALSE, xlab = "", ylab = "",
     main = "rainbow(8)")
rect(0:7, 1.1, 1:8, 1.9, col = rb, border = NA)
rect(0:7, 0.1, 1:8, 0.9, col = killer_cvd(rb, "deutan"), border = NA)
text(c(-0.2, -0.2), c(1.5, 0.5), c("normal", "deutan"), adj = 1, xpd = NA)
par(op)

min(killer_check(colours = rb)$min_distance)
```

Two of those eight colours are essentially identical to a deuteranopic reader.
Every killerPals qualitative palette keeps its worst pair above 10, and the
per-album palettes above 15:

```{r all-checks}
worst <- vapply(killer_names(), function(nm) {
  min(killer_check(nm)$min_distance)
}, numeric(1))
round(sort(worst), 1)
```

## The individual metrics

The building blocks are exported too, and work on any colours:

```{r metrics}
# Perceptual difference
killer_distance("#FF0000", "#FF3300")

# WCAG contrast ratio, and relative luminance
killer_contrast("#FF524F", "white")
killer_luminance(c("white", "grey50", "black"))

# Which pair in a palette is the weakest link?
m <- killer_distance_matrix(killer_pal("flash"), cvd = "deutan")
round(min(m, na.rm = TRUE), 1)
```

## Choosing between palettes

`killer_palette_info()` gives a tidy overview, including each palette's blurb:

```{r info}
info <- killer_palette_info(type = "qualitative")
info[order(-info$n), c("name", "album", "year", "n")][1:5, ]
```

```{r blurbs}
cat(paste0(format(info$name, width = 15), info$blurb), sep = "\n")
```
